Soybeans for the Future: Population Genetic Analysis Of A Historical Swedish Soybean Population For Future Germpasm Utalisation

### 2023-09-21 manuscript_soy.tex
21 sep

 \documentclass[9pt, onecolumn,twoside]{gsajnl}
% how i want it for inport into word: \documentclass[9pt, onecolumn,twoside]{gsajnl}
% how i want it for pdf: \documentclass[9pt,twocolumn,twoside,lineno]{gsajnl}
% Use the documentclass option 'lineno' to view line numbers

\usepackage{epstopdf}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\articletype{inv}
\runningtitle{Soy scientific manuscript} % For use in the footer
\runningauthor{Connelly \textit{et al.}}

\title{SoyAdapt}

@@

\begin{abstract}
Soybean (Glycine max L. Merrill) is the worlds leading  oilseed crop used as a  primary source of vegetable oil for human consumption and in protein meal for animal feed.

Search engine optimization (SEO)

There are a few simple ways to maximize your article's discoverability and search results.

    Include a few of your article's keywords in the title of the article

    Do not use long article titles

    Pick 5-8 keywords using a mix of generic and more specific terms on the article subject(s)

    Use the maximum amount of keywords in the first two sentences of the abstract

    Use some of the keywords in level 1 headings

    Title

The title should be concise, omitting terms that are implicit and, where possible, be a statement of the main result or conclusion presented in the manuscript. Abbreviations should be avoided within the title.

Witty or creative titles are welcome, but only if relevant and within measure. Consider if a title meant to be thought-provoking might be misinterpreted as offensive or alarming. In extreme cases, the editorial office may veto a title and propose an alternative.

Authors should avoid:

    titles that are a mere question without giving the answer

    unambitious titles, for example starting with 'Towards,' 'A description of,' 'A characterization of' or 'Preliminary study on'

    vague titles, for example starting with 'Role of', 'Link between', or 'Effect of' that do not specify the role, link, or effect

    including terms that are out of place, for example, the taxonomic affiliation apart from species name.

For Corrigenda, General Commentaries, and Editorials, the title of your manuscript should have the following format.

    'Corrigendum: [Title of original article]'

    General Commentaries:
    'Commentary: [Title of original article]'
    'Response: Commentary: [Title of original article]'

    'Editorial: [Title of Research Topic]'
\end{abstract}

\keywords{Soy; population genetics; Diversity; Selection; more keywords}

\dates{\rec{01 09, 2023} \acc{01 09, 2023}}

@@

\textit{what we already know:} Several severe genetic bottlenecks occurred in soybean. Compared to the wild species, the genetic diversity was halved, resulting in a loss of 81\% of rare alleles (Hyten et al., 2006). Additionally, two major bottlenecks occurred during the development of North American modern cultivars, where only a few landraces were used. And as a result of intensive soybean breeding over the past 75 years. Elite cultivars have emerged, but they are derived from only about 19 landraces. Consequently, the North American breeding pools now retain only 72\% of genome diversity and have lost 79\% of rare alleles found in diverse landraces (Gizlice et al., 1996).

\textit{soybean needs to be grown in the north}:
Modern soybean breeding has led to yield increases of $229 \, \text{kg} \, ha^{-1} \, \text{yr}^{-1}$ (Rincker et al., 2014), gaining a deeper understanding of the genomic basis of this progress could offer valuable insights for advancing and adapting soybean cultivation even further.

Expanding the production of soy to the northern climates by leveraging the advancements in soybean breeding and focusing on the genes relevant to adaption, can diversify the global supply chain, reduce dependency, and potentially alleviate the environmental pressures associated with soy production in high biodiversity, sensitive areas/ecologically vulnerable regions. With the global demand for soy and the need for sustainable agricultural practices, investing in breeding programs to develop soybean varieties suitable for northern climates can present a promising solution.

\textit{the Swedish soybeans (goals and scope)}

\textit{mention approaches}

\textit{In this study we}
The goal is to help precipitate the utilisation of these soybean genetic resources. to increase germplasm utilization for through understanding what we have in the genebanks.

@@

\subsection{Swedish Breeding Program Collection}
155 soybean accessions were obtained from the Nordgen genebank of Swedish origin. These accessions originate from a Swedish breeding program (SBP) running from the 1840s to the 1970s which used material consisting of a mixture of elite cultivars of the time and Japanese landraces, breeding for the adaptation to the cool climate of northwestern Europe \cite{holmberg1973}.

\subsection{The Core Collection}
The Core Collection (CCA) consists of two samplings of the core collection from haupt20. A 5\% and 10\% collection with overlapping accessions in total consisting of 409 accessions after filtering. 
The Core collection accessions (CCA) used here is a subset of the large soybean germplasm collection, the  USDA genebank accessions consisting of 415 of the USDA soy germplasm accessions. This Core collection has been selected with a focus on adaptation to high-latitude cold regions using environmental data from phenotypic trials in Germany, and comparing Donor opulation of Environments (DPE) in Asia and the Target Population of Environments (TPE) in Central Europe by \cite{haupt20}. From the more than 17.000 accessions, two diverse core collections of 183 and 366 accessions were created. These 514 diversity panels are used here due to that they are likely preadapted to cultivation in Central Europe, while simultaneously conserving a high level of genetic diversity.

\subsection{The Founders Collection}
The USDA Soybean Germplasm Collection maintains accessions of >17000 cultivated soybeans and is part of the U.S. National Plant Germplasm System (NPGS).  These accessions have been previously genotyped with the SoySNP50K iSelect BeadChip  \cite{song13,song15} has been used for genotyping the USDA Soybean Germplasm Collection and is available for use from SoyBase (https://soybase.org/) in the version mapped to the soybean genome assembly Glyma.Wm82.a2  \cite{schmutz10}. In the pedigree information from Nordgen SBPA, are 10 accessions present in the USDA genebank and thus are part of the SoySNP50K genotypic data. These specific accessions were included as they are considered Founders of the SBP.

@@

\subsection{Genotyping and filtering of whole genome sequence SNP}
The accessions from the Nordgen genebank and the Core collection accessions were whole genome sequenced. Four of the Core collection seeds didn't germinate in time for the sequencing, resulting in accessions sequenced 574. Of the Nordgen accessions, Four were of Latvian origin and therefore excluded in further analysis. 
The reads were aligned to the Williams 82 2a reference genome \cite{schmutz10} using an adapted GATK Best Practices \cite{mckenna10}, keeping MQ30 and biallelic sites and using VCFTools \cite{danecek11} applying a minor allele frequency (MAF) filter of 0.01 in order to remove rare variants from the data, reducing the number of SNPs markers were initially called (raw dataset) from 17,648,123 to 10,000,122. 
For a detailed account of accessions, programs and commands used see supplementary material: \nameref{sec:supplementary:material} 
A further three accessions were removed due to missing phenotype data and a single accession that had missing data > 5$\%$ using BCFtools \cite{danecek21}, resulting in 153 Swedish accessions from the NordGen genebank and 409 of the CCA accessions. Five per cent missing data at SNPs was also removed resulting in an SNP count of 8,533,444. This data set was used to compare the Allele Frequency of the two populations. For all other analyses, MAF 5\%  was removed using Vcftools (--maf 0.05) on the dataset as a whole, except for the case of LD calculation MAF5\% was removed separately for SBPA and CCA.

@@

@@

The package SNPRelate in R was used to compute an identity-by-state (IBS) matrix that calculates the proportion that two randomly selected reads that contain a certain SNP locus are the same or different between two individuals. The resulting pairwise IBS matrix was used to generate a cluster dendrogram using the function \textit{snpgdsHCluster: Hierarchical cluster analysis In SNPRelate }in R.

Relatedness measures F-statistics (also known as fixation indices) for given populations were calculated using the function \textit{snpgdsFst: F-statistics (fixation indices) in SNPRelate }in R. The whole genome data was used to calculate the pairwise population differentiation measured in the populations of the Swedish accessions as a subpopulation with the Core Collection as a representation of the total population. Fst is comparing the expected heterozygosity in the subpopulation (HS) to that expected in the total population (HT) :

@@

Later in the analysis when looking for signatures of selection Fst is calculated according to the methods of \cite{Weir84} using the Vcftools \cite{danecek11} function --weir-fst-pop on the SNP variants with (no window size) a window size of 1 Mb with a step size of 100 kb.

\subsection{Allele frequency, LD and nucleotide diversity}
The idea is to look into the effects of the bottleneck caused by the SBP, the increased drift and will be detected as a reduction of Ne. Detecting a decrease in genetic diversity as a measure of allele frequency, nucleotide diversity and LD increase. Allele Frequency (AF) was calculated using Vcftools --freq and plotted in R. For the LD analysis vcftools --maf 0.05 for each population was removed separately. The function \textit{snpgdsLDmat: Linkage Disequilibrium (LD) analysis, in SNPRelate } in R. A matrix of "r" for R coefficient (by EM algorithm assuming HWE) is calculated for pairwise snps separately for each chromsome. 
The squared correlation coefficient r2 is a popular measure of LD an chosen due to the advantage of this calculation being less sensitive to marginal allele frequencies over  D or D' \cite{sved18}.

@@

The $r^2$  values were binned and plotted using ggplot2 \cite{wickham16} in R (ref).  Nucleotide diversity ($\pi$), measuring the degree of polymorphism in the populations was calculated with the wgs data using vcftools for 100 kb windows --window-pi 10000 \cite{nei79}. *This is appropriate @ 25kb as we just discussed earlier today with the fst.

\subsection{H3}
Q and Method

@@

Indicate what statistical analysis has been performed.

\section{Results and Discussion}

\subsection{H1 The Swedish soybean accessions were part of the SBP}
The accessions from the NordGen gene bank (Swedish breeding program accessions, SBPA) are closely related and therefore we can conclude that they were part of the same breeding program, the Swedish breeding program (SBP), of the 1940s-1980s of Algot Holmberg and Sons from Norrköping, producing the known cultivars, Fiskeby III, Fiskeby V, Träff $\&$ Bråvalla, which are included in the NordGen gene bank accessions and in this analysis. Two accessions are identified as accessions that were not part of the Swedish breeding program of interest but have been part of other Swedish or Nordic soybean breeding trajectories and later obtained from the Russian seed bank.

\subsubsection{Relatedness: The SBPA are genetically similar and therefore closely related}
There are 155 soybean accessions sequenced from the Nordgen genebank. 153 of the accessions are similar to a degree that we can conclude they are closely related to a degree of XXX.  numbers from the ibs?
Two of the Swedish accessions (Altonagård \& Ugra Soja) show up clearly distinct from the cluster of SBPA and are therefore no longer considered or included as accessions of the investigated (SBP).  
\begin{figure}
\centering
\includegraphics[width=\linewidth]{plot_heatmap.pdf}
\caption{This heatmap includes Altonagård and Ugra soja which are seen as the bottom two blue acessions. They are dissimilar to eachother and dissimmilar to the rest of the accessions.}
\label{fig:heat}
\end{figure}

Characterise genetic relatedness. 
PCA the first 10 pcs capture 22.82\%  of the variation.
the ibs shows identity towards 1 to be at: numbers?

\subsubsection{Distinctness: The SBPA are a genetically cohesive group}
The SBPA are a genetically cohesive group of accessions. In comparison to the CCA, they are similar to each other (matrix numbers?) as seen in the figure the SBPA cluster within the dendrogram of the CCA and the SBPA.  It is also clear that some of the CCA are closely related to the SBPA, which is expected as a domesticated crop and the history known of other pedigrees and breeding programs exchanging germplasm. In the CCA, 94 of the accessions are of European origin and additionally, seven of those  originate from Sweden, some with accession names identical or similar to that of the SBPA. 
\begin{figure}
\centering
\includegraphics[width=\linewidth]{rainbow.pdf}

@@

\subsubsection{Divergence: There is a (significant) divergence within the subpopulation SBPA of the soybean when comparing other groups of soybean}

Genetic differentiation, e.g., Fst was calculated using SBPA as a subpopulation of the CCA. This is argumentatively valid since the CCA has a variability comparable to the soybean germplasm as a whole (ref). Fst mean at 0.178
The CCA has furthermore been grouped into subpopulations according to Maturation score status to compare the divergence of the SBPA. 
Figure: Fst
The high Fst of the SBPA shows that there is a considerable degree of differentiation in comparison to the other groups defined by maturity group shows

A relatively high Fst mean between (0.405 \& 0.215) for the SBPA in comparison to the CCA groups shows that there is a considerable degree of differentiation in relation to the other populations as defined by the maturity group. 
The CCA in relation to each other has a relatively low Fst, between (0.075 \& 0.007) showing that differentiation is not high when grouping the CCA by Maturity.

\textbf{Founders}
p4. summary of what we can see of the founders. 
A list of known and suspected founders has been included in the PCA. It is clear that (some specific) founders are closely related to the SBPA and were clearly founders of the SBP whereas others are not distinguishable as Founders. It is expected that the founder accessions could be more related to accessions in the CCA than the SBPA. A more detailed picture of decent can be seen in the supplementary material where the pedigree data of the SBPA are collected. \nameref{sec:supplementary:material}.

Figure: PCA with founders in. 
The founders could also be indicated earlier in the dendrogram clustering together with the SBPA and the SBPA cluster nested within the founders.

\subsection{H2 Genetic diversity}

\textbf{The SBP caused a bottleneck as seen in a decrease in genetic diversity }

There is a reduction of Ne detected in a high level of linkage of the SBPA, the lower nucleotide diversity ($\pi$ ) There is also an accumulation of rare alleles and a lot of alleles fixed in the population. These observations suggest that the bottleneck event and the ensuing intensification of genetic drift had a noticeable impact on the genetic composition of the SBP population.

>>>>

 \documentclass[9pt, twocolumn,twoside]{gsajnl}
% how i want it for inport into word: \documentclass[9pt, onecolumn,twoside]{gsajnl}
% how i want it for pdf: \documentclass[9pt,twocolumn,twoside,lineno]{gsajnl}
% Use the documentclass option 'lineno' to view line numbers

\usepackage{epstopdf}
\usepackage{wrapfig}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\articletype{inv}
\runningtitle{Soy scientific manuscript} % For use in the footer
\runningauthor{Connelly \textit{et al.}}

\title{Soybeans for the Future: Population Genetic Analysis Of A Historical Swedish Soybean Population For Future Germpasm Utalisation}

%Swedish Soybean Genetic Diversity: Influence of Breeding Practices and Identification of Selection Signatures

%Genetic Diversity in a Swedish Soybean Population: Signatures of Selection and Implications for Future Soybean Breeding in Northern Environments

% Swedish Soybean Genetic Diversity: Influence of Breeding Practices and Identification of Selection Signatures

%Soybean Genetic Variation in Swedish population: A Study of Breeding Influence and Selection Signatures

%Swedish Soybean Genetic Diversity: Influence of Breeding Practices and Identification of Selection Signatures

%Genetic Diversity in the Swedish Soybean Population: Impacts of the Breeding Program, Selection Signatures, and Implications for Future Germplasm Utilization

%Population Genetic Insights into the Swedish Soybean Population: Impact of Breeding Strategies and Selection Signatures

@@

\begin{abstract}
Soybean (Glycine max L. Merrill) is the world's leading oilseed crop used as a  primary source of vegetable oil for human consumption and protein meal for animal feed. 
A breeding program of the 1940s-1970s resulted in 160 accessions later donated to the seed bank of the Nordic Genetic Resource Center (NordGen). Climate change and plant protein independence have driven the idea of trying to repeat the effort of plant genetic enhancement for soybean cultivation in northwest Europe. Uncovering the population genetics of this as yet unutilised germplasm can give a head start to these future soybean breeding efforts. The Swedish soybeans are revealed as a population that has undergone bottlenecks, resulting in reduced effective population size (Ne), elevated levels of genetic drift, decreased nucleotide diversity, and increased linkage, all attributed to the selective breeding efforts or the selection of the genetic progenitors.

The Core collection (CCA) was created by \cite{haupt20}  as a collection identified from the USDA germplasm for central European soybean breeding. Comparison of the CCA to the Swedish breeding program accessions (SBPA) has been used throughout the paper to put the SBPA into the perspective of germplasm likely to be utilised in these future soybean breeding efforts. The SBPA has half the nucleotide diversity of the CCA. Fst analysis for population subdivision reveals a population difference of 0.22 calculated for a weighted Fst estimate.

genomic regions of selection differentiation were detected between the CCA and the SBPA by using snp specific genomewide pairwise Fst and to uncover

The reduced level of genetic variation in elite breeding populations means that in the case of soybean, there is a need to increase germplasm utilization for abiotic adaptation to the colder conditions of Northern Europe. This research means to examine the genetic structure and diversity within a population of soybeans originating from a breeding program during the 1940s to 1970s in Sweden (SBPA). We aim to gain insights into the pressures that have shaped the genetic makeup of these soybean accessions. So that this resource of untold potential can be fully utilized.

\end{abstract}

\keywords{Soybean; Glycine max; Population Genetics; Diversity; Fst; Linkage Disequalibrium; Selection Signatures; Plant Breeding}

\dates{\rec{29 09, 2023} \acc{29 09, 2023}}

@@

This is a link to part A where there is the extensive introduction/litterature review:
https://www.overleaf.com/2689273778mwyzpfmjvbmf

ps abrtract is in the preamble...

Soybean is Glycine max, this and that and important.

Several severe genetic bottlenecks occurred in soybean during domestication and the continuous breeding efforts. Compared to the wild species, the genetic diversity was halved, resulting in a loss of 81\% of rare alleles (Hyten et al., 2006). Additionally, two major bottlenecks occurred during the development of North American modern cultivars, where only a few landraces were used and as a result of intensive soybean breeding over the past 75 years. Elite cultivars have emerged, but they are derived from only about 19 landraces. Consequently, the North American breeding pools now retain only 72\% of genome diversity and have lost 79\% of rare alleles found in diverse landraces \cite{gizlice96}.

soybean needs to be grown in the north:

With the European Green Deal of 2019, there is an intention of harnessing the potential of EU plant protein production. This intention is motivated as a response to the needs of farmers, producers and consumers. 
The production and sourcing of plant-based proteins in the EU also carries with it a range of environmental benefits such as a decrease in imports and therefore transport pollution reduction.

The difficulties identified by the Report: Development of plant proteins in the European Union \cite{EC18} are:

- the agronomic conditions in Europe, which are not optimal for large-scale production of plant proteins; 
- the economic profitability of these crops in Europe; 
- the competitiveness of EU protein crops compared to imported plant proteins; 
- competition over the use of arable land; 
- and a lack of research on breeding, agronomic practices and different uses.

(Danish field trials have shown the Swedish cultivar Fiskeby V to yield only 11.2hkg/ha \cite{petersen09,petersen10}. Modern soybean breeding has led to yield increases of $229 \, \text{kg} \, ha^{-1} \, \text{yr}^{-1}$ \cite{rincker14}, advancing and adapting soybean cultivation even further. )

Expanding the production of soy to the northern climates by leveraging the advancements in soybean breeding and focusing on the genes relevant to adaptation can diversify the global supply chain, reduce dependency, and potentially alleviate the environmental pressures associated with soy production in high biodiversity, sensitive areas/ecologically vulnerable regions. With the global demand for soy and the need for sustainable agricultural practices, investing in breeding programs to develop soybean varieties suitable for northern climates can present a promising part of the solution.

\textit{the Swedish soybeans (goals and scope)}
The Swedish breeding population SBPA consists of the soybean germplasm collection from the seed bank of the Nordic Genetic Resource Center (NordGen). One of the cultivars that resulted from this breeding program is well-known as a genetic source of 
goals
not technical

scope

\textit{In this study}

In this study, we present insights into the population genetic analysis of the soybean accessions obtained from the NorGen genebank accessions. 
There haven't been any past investigations into this germplasm resource, so therefore, the goal of this project was to better understand the value of the genebank collection so as to help precipitate the utilisation of these soybean genetic resources. to increase germplasm utilization through understanding what we have in the genebanks. We present population stratification, diversity and linkage of the population. Through population genetic analyses supported by breeders publications and historical documents, we can infer what happened and what  to this population and determine the genetic significance of the resource. 
Presented is a comprehensive population genetic analysis of the SBPA soybean population. Determining the genetic diversity and population structure of the SBPA. An assessment of how the SBPA compare to the broader genetic diversity in the species through a comparative analysis with a core collection (CCA). 
And a test of selection efficacy of the Swedish breeding program by screening biological pathways.

@@

\subsection{Swedish Breeding Program Collection}
155 soybean accessions were obtained from the Nordgen genebank of Swedish origin. These accessions originate from a Swedish breeding program (SBP) running from the 1840s to the 1970s, which used material consisting of a mixture of elite cultivars of the time and Japanese/Siberian landraces, breeding for the adaptation to the cool climate of northwestern Europe \cite{holmberg1973}. The Swedish breeding program (SBP) of the 1940s-1980s of Algot Holmberg and Sons from Norrkoping produced the known cultivars Fiskeby III, Fiskeby V, Träff, and Bråvalla, which are included in the NordGen gene bank accessions and in this analysis excluding Fiskeby III that was not available from the Nordgen genebank at the time.

\subsection{The Core Collection}
The Core Collection (CCA) consists of two samplings of the core collection from haupt20. A 5\% and 10\% collection with overlapping accessions in total consisting of 409 accessions after filtering. 
The Core collection accessions (CCA) used here is a subset of the large soybean germplasm collection, the  USDA genebank accessions consisting of 415 of the USDA soy germplasm accessions. This Core collection has been selected with a focus on adaptation to high-latitude cold regions using environmental data from phenotypic trials in Germany and comparing Donor opulation of Environments (DPE) in Asia and the Target Population of Environments (TPE) in Central Europe by \cite{haupt20}. From the more than 17.000 accessions, two diverse core collections of 183 and 366 accessions were created. These 514 diversity panels are used here due to that they are likely preadapted to cultivation in Central Europe while simultaneously conserving a high level of genetic diversity.

\subsection{The Founders Collection}
The USDA Soybean Germplasm Collection maintains accessions of >17000 cultivated soybeans and is part of the U.S. National Plant Germplasm System (NPGS).  These accessions have been previously genotyped with the SoySNP50K iSelect BeadChip  \cite{song13,song15} has been used for genotyping the USDA Soybean Germplasm Collection and is available for use from SoyBase (https://soybase.org/) in the version mapped to the soybean genome assembly Glyma.Wm82.a2  \cite{schmutz10}. In the pedigree information from Nordgen SBPA, there are 10 accessions present in the USDA genebank and thus are part of the SoySNP50K genotypic data. These specific accessions were included as they are considered Founders of the SBP.

@@

\subsection{Genotyping and filtering of whole genome sequence SNP}
The accessions from the Nordgen genebank and the Core collection accessions were whole genome sequenced. Four of the Core collection seeds didn't germinate in time for the sequencing, resulting in accessions sequenced 574. Of the Nordgen accessions, Four were of Latvian origin and therefore excluded in further analysis. 
The reads were aligned to the Williams 82 2a reference genome \cite{schmutz10} using an adapted GATK Best Practices \cite{mckenna10}, keeping MQ30 and biallelic sites and using VCFtools v. 0.1.16 \cite{danecek11} applying a minor allele frequency (MAF) filter of 0.01 in order to remove rare variants from the data, reducing the number of SNPs markers were initially called (raw dataset) from 17,648,123 to 10,000,122. 
For a detailed account of accessions, programs and commands used see supplementary material: \nameref{sec:supplementary:material} 
A further three accessions were removed due to missing phenotype data and a single accession that had missing data > 5$\%$ using BCFtools \cite{danecek21}, resulting in 153 Swedish accessions from the NordGen genebank and 409 of the CCA accessions. Five per cent missing data at SNPs was also removed resulting in an SNP count of 8,533,444. This data set was used to compare the Allele Frequency of the two populations. For all other analyses, MAF 5\%  was removed using VCFtools (--maf 0.05) on the dataset as a whole, except for the case of LD calculation MAF5\% was removed separately for SBPA and CCA.

@@

* here insert a table with information of the snps and stuff.

@@

The package SNPRelate in R was used to compute an identity-by-state (IBS) matrix that calculates the proportion that two randomly selected reads that contain a certain SNP locus are the same or different between two individuals. The resulting pairwise IBS matrix was used to conduct a hierarchical cluster analysis to generate a dendrogram using the function \textit{snpgdsHCluster: Hierarchical cluster analysis In SNPRelate }in R.

F-statistics, particularly Fst, quantifies the distribution of genetic diversity by measuring the variation in allele frequencies among different populations. 
Relatedness measures F-statistics (also known as fixation indices) for given populations were calculated using the function \textit{snpgdsFst: F-statistics (fixation indices) in SNPRelate }in R. The whole genome data was used to calculate the pairwise population differentiation measured in the populations of the Swedish accessions as a subpopulation with the Core Collection as a representation of the total population. Fst is comparing the expected heterozygosity in the subpopulation (HS) to that expected in the total population (HT) :

@@

\subsection{Genetic Diversity Estimates}
The idea is to look into the effects of the bottleneck caused by the SBP, the increased drift and will be detected as a reduction of Ne. Detecting a decrease in genetic diversity as a measure of allele frequency, nucleotide diversity and LD increase. Allele Frequency (AF) was calculated using Vcftools --freq and plotted in R. LD analysis was preformed in VCFtools  --maf 0.05 for each population was removed separately. The function \textit{snpgdsLDmat: Linkage Disequilibrium (LD) analysis, in SNPRelate }in R. A matrix of "r" for R coefficient (by EM algorithm assuming HWE) is calculated for pairwise SNPs separately for each chromosome. 
The squared correlation coefficient r2 is a popular measure of LD an chosen due to the advantage of this calculation being less sensitive to marginal allele frequencies over  D or D' \cite{sved18}.

@@

The $r^2$  values were binned and plotted using ggplot2 \cite{wickham16} in R (ref).  Nucleotide diversity ($\pi$), measuring the degree of polymorphism in the populations, was calculated with the wgs data using VCFtools for 100 kb windows --window-pi 10000 \cite{nei79}.

\subsection{Selection Signature Detection }
For testing signatures of selection, TajimasD statistic and pairwise FstFst is calculated according to the methods of \cite{Weir84} using the Vcftools function --weir-fst-pop on the SNP variants with no sliding window, but a window size of 25kbp --fst-window-size 25000. The same window is used for calculating TajimasD, --TajimaD 25000. This window-based approach for Fst statistics has proven successful in minimizing sampling error while preserving the genuine signal when investigating evidence of selection in populations \cite{beissinger15}.

Candidate gene enrichment analysis. 
Pairwise Fst values across the same window were used to assess the divergence between the SBPA and the CCA and Tajima's D for each population was calculated over the same window. 
gff3 annotation of the genome data was used to further analyse whether the higher Fst values or low Tajima's D values are significant for genes correlated to photoperiod or cold tolerance traits. 
Pathway screening analysis.
To look for enrichment or degradation of genes in other pathways enrichment was tested.

GeneIds, locations and GO functions were downloaded via Soybase.org and genes were selected based on a word search of the Gene Ontology biological process descriptions for flowering and cold tolerance. A further 14 genes were selected for enrichment analysis based on genes known to be involved in photoperiod regulation.   
Enrichment analysis was used to see if there was a significant difference by testing the significance of the number of genes above or below an Fst and TajimasD threshold. For the Fst threshold, the upper five per cent value was used and for Tajimas D, any value under zero. For details on the analysis, see the link to the R code in the section on data availability.  \label{sec:data:availability}

@@

Fisher's exact test was chosen to test significance of enrichment because Chi squared can be unreliable if some contingency tables have cells of low numbers. This is the case in the pathway screening analysis, and alternatively the Fisher's exact test can be used when there are low expected values. The test gives the exact probability of the observed values \cite{quicke21}.  Fisher's exact test was used to test for the pathway screening analysis (the data are counts comparing variables) to test whether there is a disparity between the variables, frequencies of genes above and below the chosen threshold of Fst and TajimasD of genes within or not within the pathway genes.

Caveats of the statistical test. 
Replicates? Fst is calculated on the 160 samples of The SBPA as a population. but if there is a very high degree of similarity in the population, there are effective pseudo-replicates of the samples.

\section{Results}

The Swedish accessions are of a cohesive unit, the same breeding program; they are genetically similar, and there are clear indications that the population has undergone a recent bottleneck, that leads us to infer that the is a small effective population size (Ne). Effective population size determines genetic variation, genetic drift and linkage disequilibrium in populations \cite{tenesa07}. The results of these effects are seen to indicate drift and low genetic variation, a high number of fixed alleles and high LD decay. 
Despite strong drift, a screening of the SBPA has resulted in evidence of enrichment in 34 pathways and evidence of a degradation in three pathways.

\subsubsection{Relatedness: The SBPA are genetically similar and, therefore closely related}
The SBPA are closely related, and therefore, we can conclude that they were part of the same breeding program. This is supported by three sources of metadata: firstly, the names of the accessions align with the breeders' coding of crosses; secondly, the seeds were contributed by the breeding company; and thirdly, the pedigree information further substantiates this finding \cite{holmberg1973}.

In the initial analysis, we discovered two accessions, namely Altonagård and Ugra Soja, which, while not originally associated with the specific Swedish breeding program of interest, have significant historical ties to other Swedish and Nordic soybean breeding initiatives. These accessions were subsequently acquired by the Nordgen genebank through a Russian genebank. It's worth noting that these two accessions do not align with the SBPA in either the hierarchical cluster dendrogram where we consider pairwise identity-by-state (IBS) analysis or the PCA analysis, as illustrated in a PCA of the accessions aquired from the Nordgen genebank, see \nameref{sec:supplementary:material} S5.1

Figure \ref{fig:pca}. shows two main clusters, the SBPA clustered together and the larger CCA clustered above. The accessions identified as the Founders of the breeding program are situated between and within the two clusters. Figure \ref{fig:pca}.  shows the first two principal components (PCs) of the PCA, comprising 10.3 and 3.4\% of the variation, respectively. The PCA captures 22.82\% of the variation in the first 10 PCs where combinations of the first 4 can be seen in the Supplementary Material figure S5.2. It is clear that the SBPA clusters together on all 6 combinations of the first 4 PCs. 
The high sample number of the SBPA (153) along with the high similarity drags the SBPA cluster away from the larger cluster consisting of the CCA. The importance of the sample size for distance in PC1 is demonstrated in S5.2 where a smaller sample size of the SBPA is used. Here it is clear that although the SBPA accessions are not in a cluster at a distance from the rest, the SBPA still cluster together in all of the first 4 PCs explaining together 12.8\% of the variance, confirming the SBPA similarity.

The Hierarchical cluster analysis based on pairwise identity-by-state (IBS) values from the wgs SNPs also indicates the SBPA as a group with high similarity.  *more

\subsubsection{Distinctness: The SBPA are a genetically cohesive group}
The SBPA are a genetically cohesive group of accessions, as they cluster together and are removed from the CCA as a whole. Figure IBS heatmap shows the SBPA clustered together in the top left-hand corner of the heatmap. The clustering is made from the IBS calculated similarity, and the hierarchical cluster analysis shows the SBPA clustering together within the group. This clustering has a very high similarity to the  pedigree dendrogram S2.3. 
Figure PCA shows the SBPA as clustered together by the first principal component, capturing 10.3\% of the variation.

It is also clear that there are accessions of the CCA that are closely related to the SBPA,. This is expected from a domesticated crop such as soybean because of the known pedigrees and breeding programs exchanging germplasm.  In the CCA, 94 of the accessions are of European origin. Additionally, seven of those originate from Sweden, some with accession names are identical or similar to that of the SBPA. See S2 for all accessions and  S5.1 for a PCA showing accessions with names in both datasets. A few of the CCA accessions are also thought to be a Founder of the SBPA.

\subsubsection{Divergence: There is a divergence within the subpopulation SBPA of the soybean when comparing other groups of soybean}
Population subdivision analysis will quantifys the SPBA as a distinct subpopulation with a distinct allele frequency difference.

The result of divergent selection and drift, causes genetic differentiation in separate populations over time.  
The SBPA are genetically different to the CCA, Fst was calculated using SBPA as a subpopulation of the CCA. The CCA has a variability comparable to the soybean germplasm as a whole \cite{haupt20}. 
The Weighted Fst estimate is  0.22, and the Mean Fst 0.11 between SBPA as a subpopulation of CCA, indicating a high level of genetic differentiation between the populations (Figure FST1).  
percent reduction in heterozygosity in the total population due to population structuring

To further quantify population differentiation, the CCA was grouped into subpopulations based on Maturation score status to compare the divergence of the SBPA where no maturation scores have been given Table  \ref{tab:fst}. 
The high Fst of the SBPA shows a considerable degree of differentiation compared to the other groups defined by the maturity group. A relatively high Fst mean between (0.405 \& 0.215) for the SBPA compared to the CCA groups shows a considerable degree of differentiation in relation to the other populations as defined by the maturity group. The CCA in relation to each other has a relatively low Fst, between (0.075 \& 0.007), showing that differentiation is not high when grouping the CCA by Maturity.

\subsubsection{Founders}
A list of known and suspected founders has been included in the figure PCA. It is clear that some of the founders are closely related to the SBPA as they are situated between the SBPA and CCA clusters. This corroborates the idea of these accessions being founders of the SBP. Additionally, accessions of the founders cluster on the IBS hierarchical analysis close to the SBPA (S5.x). They are also in the PCs situated close to the SBPA. Other founder accessions are not distinguishable as Founders in this way. There are also CCA accessions that have been found to actually be SBPA. all this is noted in the large IBS dendrogram of the accessions and in the S2 accessions names and annotations  It is also expected that a founder accession could be more related to accessions in the CCA than the SBPA because of these accessions use in other breeding tajectories. A more detailed picture of decent can be seen in the supplementary material where the pedigree data of the SBPA are collected. \nameref{sec:supplementary:material}. S2.3.

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{PCA1.pdf}
\caption{Biplot of the principal component analysis of soybean genotypes. The colours represent the defined origin groups, and the shapes indicate whether the accessions are from the Core Collection Accessions (CCA), the Swedish Breeding Program Accessions (SBPA) or the Founders of the breeding program (F).}
\label{fig:pca}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{heat1.pdf}
\caption{This heatmap is a visual representation of the IBS values of the SBPA and CCA accessions. IBS values that are from 0 to 1 where 1 is identical by state. In the upper left corner are the SBPA accessions collected by the IBS matrix and hierarchical clustering-based analysis. The SBPA IBS values range from 0.73 to 0.97 for the SBPA and 0.56 to 0.98 for the CCA. The dataset as a whole, therefore ranges from 0.56 to 0.98.  }
\label{fig:heatall}
\end{figure}
\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{rainbow.pdf}

@@

\begin{wraptable}{r}{0.5\textwidth}
\begin{table}[p]
\centering 
\caption{$F_{\text{ST}}$ estimation on genotypes. Measuring the population differentiation of  the 155 Accessions of the Nordgen genebank (Do again with only the 153 that were part of the breeding program)(SBPA) to the Core Collection of 409 accessions (CCA).}

Samples:                        &564        &          &          &          &          &          \\ \hline
SNPs:                           & 8,533,444 &          &          &          &          &          \\ \hline
Method: Weir \& Cockerham, 1984 &           &          &          &          &          &          \\ \hline
Populations:                  & CCA       & 409      & SBPA     & 155      &          &          \\ \hline
Weighted Fst estimate:          &0.2231884  &          &          &          &          &          \\ \hline
Mean Fst:                       &0.1108533  &          &          &          &          &          \\ \hline
summary:                        & Min.      & 1st Qu   & Median   & Mean     & 3rd Qu.  & Max.     \\ \hline
                                & -0.002515 & 0.011669 & 0.041730 & 0.110853 & 0.145697 & 0.975485 \\ \hline
\end{tabular}
  \label{tab:fst}
\end{tableminipage}
\end{table}
\end{wraptable}
\begin{wraptable}{r}{0.5\textwidth}
\begin{table}
\centering 
\caption{$F_{\text{ST}}$ estimation on Maturity groups, measuring the population differentiation of  the 155 Accessions of the Nordgen genebank that were part of the breeding program)(SBPA) to the Core Collection of 409 accessions (CCA) groups by the corrosponding maturity groups.}
